Dialect: A Compact Line Notation Based on SMILES

### 2022-03-28 manuscript.tex
Improves the flow of Constitution and Bond Elision.

A Dialect molecule is described at the lowest level by its constitution. Constitution consists of a set of atomic nuclei (nodes), a set of pairwise bonding relationships between them (edges), and those attributes needed to associate every valence electron with a specific node or edge. Constitution excludes attributes related to stereochemistry, or any other feature whose attributes extend over more than one atom or bond.

Two attributes characterize an atom's nucleus: \ttt{element} and \ttt{isotope}. The `element` attribute is an optional one- or two-letter character sequence selected from the set designated by the IUPAC~\cite{periodicTable}. When IUPAC authorizes additional symbols, they will become valid values for the \ttt{element} attribute. If the elemental identity of an atom is unknown, its \ttt{element} attribute can be omitted. The \ttt{isotope} attribute is an optional integer value representing an atom's nuclear mass number, where mass number is the sum of proton and neutron count. Omitting the \ttt{isotope} property means that the isotopic composition is unspecified. When present, the lower bound on \ttt{isotope} equals one. This constraint leaves room for physically meaningless negative implied neutron count as in, for example, carbon-5. The lower bound for isotope applies even if the \ttt{element} attribute is not defined.

The \ttt{hydrogens} attribute encodes an atom's mode of \textit{hydrogen suppression}. Hydrogen suppression replaces hydrogen atoms in a molecular graph with a more compact representation. Despite the removal of its node, a suppressed hydrogen and its bonds to the parent atom are understood to be present. The abundance and regular bonding of hydrogen atoms in molecular graphs makes hydrogen suppression a popular strategy for increasing information density.

Dialect supports two mutually-exclusive modes of hydrogen suppression. The first is \textit{virtual hydrogen}. A virtual hydrogen is one whose node and connecting bond have been replaced by an integer tally attribute on the parent atom. For example, methane can be represented by a molecular graph having five nodes and four edges. But methane can also be represented as a graph of one node whose \ttt{hydrogens} attribute equals four. When using this variant, the \ttt{hydrogens} attribute may assume integer values from zero to nine, inclusive. The process of transforming one or more hydrogen nodes into a virtual hydrogen is called \enquote{virtualization.} The opposite process, transforming one or more virtual hydrogens into a node, is called \enquote{reification.}

The second mode of suppressed hydrogen is \textit{implicit hydrogen}. An implicit hydrogen is similar to a virtual hydrogen in that both the hydrogen node and bond have been replaced with a tally. Unlike a virtual hydrogen, however, this tally is not present as attribute on the parent atom but rather as a computation. The computation will be described in detail in a subsequent section. Subject to certain limitations, an an atom identifies itself as having an implicit hydrogen count by setting its \ttt{hydrogens} attribute to \ttt{Implicit}.

The remaining constitutional atomic attribute, \ttt{charge}, enables electron counting. Electron counting associates each electron in a molecular graph with either an atom or a bond. This is a fundamental requirement for the principle of mass conversion. In a more practical sense, electron counting lays the foundation for calculations of formal charge and bond order~\cite{dietz:1995}.

Electron counting in Dialect is based on the well-known \textit{valence bond model} (VB Model)~\cite{lewis:1916}. Although not commonly considered as such, the VB model can be thought of as an algorithm for molecular assembly. The perspective is especially useful given that Dialect itself can be viewed as a programming language for molecular assembly.

A simplified algorithm for electron counting can be summarized as follows. A molecular graph starts as a set of atoms, where each atom consists of a nucleus (as described above) and an integer electron count equal to the atomic number of the atom's element. Two nodes (nuclei) are selected as the terminals of a bond. From each atom, the same positive electron count is deducted. The total electron count deducted from each atom is simultaneously credited to the bond's electron count. For example, if each atom contributes one electron to a bond, then the bond's electron count will increment by two.

@@

From these rules follow some practical consequences. For one, bond order must be an integer greater than zero. This follows from the definition of bond formation. The electron count deducted from each atom is $n$, an integer greater than zero. Given bond formation deducts an equal number of electrons from each atom, the total number of electrons associated with the newly-formed bond is $2n$. By definition, bond order is the bond's electron count divided by two, therefore $n$ is the bond order. Fractional, negative, and zero bond orders are all disallowed by Dialect for this reason.

Dialect imposes an upper limit of three on formal bond order. The relative scarcity of bond orders greater than three in molecules conforming the the VB Model makes this restriction unlikely to meaningfully restrict applicability.

@@

@@

To promote compact string representations, Dialect supports \textit{bond elision}. A bond of order one may be setting its \ttt{elided} attribute to \ttt{true}. As such, an elided bond always has a bond order of one. A bond may be elided for one of two reasons. First, elided bonds are not written during serialization, allowing for increased information density. The second reason relates to a feature described in the next section.

>>>>

A molecular graph is described at the lowest level by its constitution. In Dialect, constitution comprises the molecular graph and those attributes locating nucleons and electrons at specific atoms and bonds. All other attributes lie outside the scope of constitution. Atomic attributes related to constitution are: \ttt{index}; \ttt{element}; \ttt{hydrogens}; and \ttt{charge}. Bond attributes related to constitution are \ttt{source}, \ttt{target}, \ttt{order}, and \ttt{elided}.

The attributes \ttt{element} and \ttt{isotope} characterize an atom's nucleus. The \ttt{element} attribute is an optional one- or two-letter character sequence selected from the set designated by the IUPAC~\cite{periodicTable}. When IUPAC authorizes additional symbols, they will become valid values for the \ttt{element} attribute. If the elemental identity of an atom is unknown, its \ttt{element} attribute must equal \ttt{None}. The \ttt{isotope} attribute is an optional integer value representing an atom's nuclear mass number, where mass number is the sum of proton and neutron count. Setting the \ttt{isotope} property to \ttt{None} means that the element's isotopic composition equals natural abundance. Otherwise, the lower bound on the value of \ttt{isotope} is zero. This lower limit allows for such physically nonsensical states such as atoms of zero implied mass (\ttt{isotope} equals zero) or negative implied neutron count (e.g., carbon-5). Implementations may reject such atomic states as invalid.

The \ttt{hydrogens} attribute sets an atom's mode of \textit{hydrogen suppression}. Hydrogen suppression replaces an atomic hydrogen of natural isotopic abundance and its associated bond with an integer tally associated with the parent atom. Even so, a suppressed hydrogen and its bond are understood to be present. The relative abundance and regular bonding of hydrogen atoms in molecular graphs makes hydrogen suppression an effective technique for increasing information density. Two mutually-exclusive types of suppressed hydrogen are supported: \textit{virtual hydrogen} and \textit{implicit hydrogen}.

A virtual hydrogen is one that appears as an integer unit contribution associated with the parent atom. For example, methane can be represented by a molecular graph having five nodes and four edges. But methane can also be represented as a graph of one node whose \ttt{hydrogens} attribute equals four. When using this mode, the \ttt{hydrogens} attribute may assume integer values from zero to nine, inclusive. The process of replacing atomic hydrogens with virtual hydrogen is called \enquote{virtualization.} The opposite process, replacing virtual hydrogen with atomic hydrogen, is called \enquote{reification.}

An implicit hydrogen is similar to a virtual hydrogen in that both the hydrogen node and bond have been replaced with a tally. Unlike a virtual hydrogen, however, this tally is not present as attribute on the parent atom but rather as a computation. Subject to certain limitations, an atom identifies itself as having an implicit hydrogen count by setting its \ttt{hydrogens} attribute to \ttt{Implicit}. The computation and its limitations will be described later in this paper.

The atomic attribute, \ttt{charge}, enables electron counting. The purpose of electron counting is to associate each electron in a molecular graph with either an atom or a bond. This is a fundamental requirement for the principle of mass conversion. In a more practical sense, electron counting lays the foundation for calculations of formal charge and bond order~\cite{dietz:1995}.

Electron counting in Dialect is based on the well-known \textit{valence bond model} (VB Model)~\cite{lewis:1916}. Although not commonly considered as such, the VB model can be thought of as an algorithm for molecular assembly. This perspective is especially useful given that Dialect itself can be viewed as a programming language for molecular assembly.

A simplified algorithm for electron counting can be summarized as follows. A molecular graph starts as a set of disconnected atoms and an integer electron count equal to the atomic number of the atom's element. Two nodes are selected as the terminals of a bond. From each atom, the same positive electron count is deducted. The total electron count deducted from each atom is simultaneously credited to the bond's electron count. For example, if each atom contributes one electron to a bond, then the bond's electron count will increase by two.

@@

From these rules follow some practical consequences. For one, bond order must be an integer greater than zero. This follows from the definition of bond formation. The electron count deducted from each atom is $n$, an integer greater than zero. Given that bond formation deducts an equal number of electrons from each atom, the total number of electrons associated with the newly-formed bond is $2n$. By definition, bond order is the bond's electron count divided by two, therefore $n$ is the bond order. Fractional, negative, and zero bond orders are all disallowed by Dialect for this reason.

Dialect imposes an upper limit of three on formal bond order. The relative scarcity of bond orders greater than this limit in molecules conforming the the VB Model makes this restriction unlikely to meaningfully restrict applicability.

@@

Each atom carries an \ttt{index} attribute, allowing bonds to reference atoms in a molecular graph. To do so, the \ttt{source} and \ttt{target} attributes are set with the respective indexes of the source and target atoms. The set of all bonds and the nodes they reference determines the connectivity of a molecular graph. In the context of determining connectivity, source and target indexes may be reversed.

@@

A bond whose attributes are assigned default values can participate in \textit{bond elision}. In bond elision, a bond's \ttt{elided} attribute is set to \ttt{true}. When written in string form, elided bonds require no characters. This makes bond elision useful for data compression. The second application for elided bonds will be described in the next section.
